My title

### 2024-11-04 paper/paper.qmd
updated paper

library(tidyverse)
library(caret)
library(glmnet)
library(nnet)
library(lubridate)
library(pROC)
library(stringr)
library(gt)
```

@@

The aggregated polling data includes support percentages of different presidential candidates in polls with different FiveThirtyEight ratings [@5382] across two elections. Please view [UPDATE APPENDIX] for summary statistics of sub-datasets.

@@

To achieve the first goal, the polls dataset was divided by individual states, and only data points from the seven swing states were selected, forming seven smaller datasets. Then, for each state, the harris_win variable was constructed by comparing the support rate of the candidates in each poll. The variable was set to 1 if the support rate of Harris is greater than or equal to Trump's support rate and 0 when Trump's support rate is greater. A logistic model was constructed for each state, using harris_win as the outcome and previously stated predictors as inputs. [EXPLAIN THE PROBABILITY STEP HERE]

@@

Our results are summarized in @tbl-modelresults.

```{r}
# plot model predictions
# combine the datasets to create a single visual
combined_data_adj <- bind_rows(harris_2024_national_adj, trump_2024_national_adj)
combined_data <- bind_rows(harris_2024_national, trump_2024_national)

# create the plot that contains scattered points of adjusted percentages
# and lines for fitted values
ggplot(combined_data_adj, aes(x = end_date, y = pct_adj, color = candidate_name)) +
  # Scatter points for actual adjusted percentages
  geom_point(alpha = 0.6) +
  # Regression lines
  geom_line(aes(y = fitted_date), size = 0.5) +
  xlim(as.Date("2024-07-19"), as.Date("2024-10-25")) +
  # Assign specific colors
  scale_color_manual(values = 
                       c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  # Customize the plot appearance
  theme_classic() +
  labs(
    y = "Adjusted Percent Support",
    x = "End Date",
    title = "Multiple Linear Models for Harris and Trump (2024)",
    color = "Candidate"
  ) +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

# Unadjusted support percentages
ggplot(combined_data, aes(x = end_date, y = pct, color = candidate_name)) +
  # Scatter points for actual adjusted percentages
  geom_point(alpha = 0.6) +
  # Regression lines
  geom_line(aes(y = fitted_date), size = 0.5) +
  xlim(as.Date("2024-07-19"), as.Date("2024-10-25")) +
  # Assign specific colors
  scale_color_manual(values = 
                       c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  # Customize the plot appearance
  theme_classic() +
  labs(
    y = "Adjusted Percent Support",
    x = "End Date",
    title = "Multiple Linear Models for Harris and Trump (2024)",
    color = "Candidate"
  ) +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

@@

```{r}
summary(model_logistic)
summary(model_logistic_adj)
```

@@

Weaknesses and next steps should also be included.

@@

## Diagnostics

```{r}
#| label: fig-pen
#| fig-cap: ROC Curve for Harris Win Prediction, Pennsylvania
#| echo: false
#| message: false
#| warning: false

@@

>>>>

library(tidyverse)
library(caret)
library(glmnet)
library(nnet)
library(lubridate)
library(pROC)
library(stringr)
library(gt)
library(broom)
library(knitr)
```

@@

The aggregated polling data includes support percentages of different presidential candidates in polls with different FiveThirtyEight ratings [@5382] across two elections.

@@

To achieve the first goal, the polls dataset was divided by individual states, and only data points from the seven swing states were selected, forming seven smaller datasets. Then, for each state, the harris_win variable was constructed by comparing the support rate of the candidates in each poll. The variable was set to 1 if the support rate of Harris is greater than or equal to Trump's support rate and 0 when Trump's support rate is greater. A logistic model was constructed for each state, using harris_win as the outcome and previously stated predictors as inputs. [EXPLAIN THE 1000 simulations and PROBABILITY calculation STEP HERE]

@@

## Swing states

@tbl-result shows the average likelihood percentage of Harris winning each swing state, calculated by using the logistic model on 1,000 hypothetical data points for each state, generated with the normal distribution of predictors and setting the end_date to election day. It indicates that Harris has a greater probability to win in Pennsylvania, Georgia, and Nevada, while Trump is more likely to win in North Carolina, Michigan, Arizona, and Wisconsin. The electoral votes in each state is also presented.

#### Results ####
# To predict the outcome on election day, generate 1000 hypothetical data points
# using the election day as end_date and the normal curve of other input 
# variables, mimicing all the situations that may happen on election day
hypothetical_data_Pennsylvania <- tibble(
  end_date = as.Date("2024-11-05"),
  has_sponsor = rbinom(1000, size = 1, prob = mean(combined_Pennsylvania$has_sponsor)),
  transparency_score = rnorm(1000, mean = mean(combined_Pennsylvania$transparency_score), 
                             sd = sd(combined_Pennsylvania$transparency_score)),
  sample_size = rnorm(1000, mean = mean(combined_Pennsylvania$sample_size),
                      sd = sd(combined_Pennsylvania$sample_size))
)

hypothetical_data_Georgia <- tibble(
  end_date = as.Date("2024-11-05"),
  has_sponsor = rbinom(1000, size = 1, prob = mean(combined_Georgia$has_sponsor)),
  transparency_score = rnorm(1000, mean = mean(combined_Georgia$transparency_score), 
                             sd = sd(combined_Georgia$transparency_score)),
  sample_size = rnorm(1000, mean = mean(combined_Georgia$sample_size),
                      sd = sd(combined_Georgia$sample_size))
)

hypothetical_data_North_Carolina <- tibble(
  end_date = as.Date("2024-11-05"),
  has_sponsor = rbinom(1000, size = 1, prob = mean(combined_North_Carolina$has_sponsor)),
  transparency_score = rnorm(1000, mean = mean(combined_North_Carolina$transparency_score), 
                             sd = sd(combined_North_Carolina$transparency_score)),
  sample_size = rnorm(1000, mean = mean(combined_North_Carolina$sample_size),
                      sd = sd(combined_North_Carolina$sample_size))
)

hypothetical_data_Michigan <- tibble(
  end_date = as.Date("2024-11-05"),
  has_sponsor = rbinom(1000, size = 1, prob = mean(combined_Michigan$has_sponsor)),
  transparency_score = rnorm(1000, mean = mean(combined_Michigan$transparency_score), 
                             sd = sd(combined_Michigan$transparency_score)),
  sample_size = rnorm(1000, mean = mean(combined_Michigan$sample_size),
                      sd = sd(combined_Michigan$sample_size))
)

hypothetical_data_Arizona <- tibble(
  end_date = as.Date("2024-11-05"),
  has_sponsor = rbinom(1000, size = 1, prob = mean(combined_Arizona$has_sponsor)),
  transparency_score = rnorm(1000, mean = mean(combined_Arizona$transparency_score), 
                             sd = sd(combined_Arizona$transparency_score)),
  sample_size = rnorm(1000, mean = mean(combined_Arizona$sample_size),
                      sd = sd(combined_Arizona$sample_size))
)

hypothetical_data_Wisconsin <- tibble(
  end_date = as.Date("2024-11-05"),
  has_sponsor = rbinom(1000, size = 1, prob = mean(combined_Wisconsin$has_sponsor)),
  transparency_score = rnorm(1000, mean = mean(combined_Wisconsin$transparency_score), 
                             sd = sd(combined_Wisconsin$transparency_score)),
  sample_size = rnorm(1000, mean = mean(combined_Wisconsin$sample_size),
                      sd = sd(combined_Wisconsin$sample_size))
)

hypothetical_data_Nevada <- tibble(
  end_date = as.Date("2024-11-05"),
  has_sponsor = rbinom(1000, size = 1, prob = mean(combined_Nevada$has_sponsor)),
  transparency_score = rnorm(1000, mean = mean(combined_Nevada$transparency_score), 
                             sd = sd(combined_Nevada$transparency_score)),
  sample_size = rnorm(1000, mean = mean(combined_Nevada$sample_size),
                      sd = sd(combined_Nevada$sample_size))
)

# run the corresponding logistic regression on the hypothetical data of each
# state and average the outcomes to find the expected win rate of Harris
# on election day
hypothetical_data_Pennsylvania <- hypothetical_data_Pennsylvania %>%
  mutate(
    harris_win_prob = round(100*predict(logistic_Pennsylvania, newdata = ., type = "response"), 2)
  )

hypothetical_data_Georgia <- hypothetical_data_Georgia %>%
  mutate(
    harris_win_prob = round(100*predict(logistic_Georgia, newdata = ., type = "response"), 2)
  )

hypothetical_data_North_Carolina <- hypothetical_data_North_Carolina %>%
  mutate(
    harris_win_prob = round(100*predict(logistic_North_Carolina, newdata = ., type = "response"), 2)
  )

hypothetical_data_Michigan <- hypothetical_data_Michigan %>%
  mutate(
    harris_win_prob = round(100*predict(logistic_Michigan, newdata = ., type = "response"), 2)
  )

hypothetical_data_Arizona <- hypothetical_data_Arizona %>%
  mutate(
    harris_win_prob = round(100*predict(logistic_Arizona, newdata = ., type = "response"), 2)
  )

hypothetical_data_Wisconsin <- hypothetical_data_Wisconsin %>%
  mutate(
    harris_win_prob = round(100*predict(logistic_Wisconsin, newdata = ., type = "response"), 2)
  )

hypothetical_data_Nevada <- hypothetical_data_Nevada %>%
  mutate(
    harris_win_prob = round(100*predict(logistic_Nevada, newdata = ., type = "response"), 2)
  )

# produce a table of mean winning probabilities
mean_probabilities <- tibble(
  State = c("Pennsylvania", "Georgia", "North Carolina", "Michigan", "Arizona", "Wisconsin", "Nevada"),
  Harris_Win_Probability = c(
    round(mean(hypothetical_data_Pennsylvania$harris_win_prob, na.rm = TRUE), 2),
    round(mean(hypothetical_data_Georgia$harris_win_prob, na.rm = TRUE), 2),
    round(mean(hypothetical_data_North_Carolina$harris_win_prob, na.rm = TRUE), 2),
    round(mean(hypothetical_data_Michigan$harris_win_prob, na.rm = TRUE), 2),
    round(mean(hypothetical_data_Arizona$harris_win_prob, na.rm = TRUE), 2),
    round(mean(hypothetical_data_Wisconsin$harris_win_prob, na.rm = TRUE), 2),
    round(mean(hypothetical_data_Nevada$harris_win_prob, na.rm = TRUE), 2)
  )
)

# include electoral votes to determine who is in the lead
state_evs <- tibble::tribble(
  ~State, ~electoral_votes,
  "Alaska", 3,
  "Arizona", 11,
  "Arkansas", 6,
  "California", 55,
  "Colorado", 9,
  "Connecticut", 7,
  "Florida", 29,
  "Georgia", 16,
  "Illinois", 20,
  "Indiana", 11,
  "Iowa", 6,
  "Kansas", 6,
  "Louisiana", 8,
  "Maine", 2,
  "Maine CD-1", 1,
  "Maine CD-2", 1,
  "Maryland", 10,
  "Massachusetts", 11,
  "Michigan", 16,
  "Minnesota", 10,
  "Mississippi", 6,
  "Missouri", 10,
  "Montana", 3,
  "Nebraska", 2,
  "Nebraska CD-2", 1,
  "Nevada", 6,
  "New Hampshire", 4,
  "New Mexico", 5,
  "New York", 29,
  "North Carolina", 15,
  "Ohio", 18,
  "Oklahoma", 7,
  "Oregon", 7,
  "Pennsylvania", 20,
  "Rhode Island", 4,
  "South Carolina", 9,
  "South Dakota", 3,
  "Texas", 38,
  "Utah", 6,
  "Vermont", 3,
  "Virginia", 13,
  "Washington", 12,
  "Wisconsin", 10
)

mean_probabilities <- mean_probabilities %>%
  left_join(state_evs, by = "State")

#### Overall Probability of Winning ####
# Harris
# Create data frame of states
states <- data.frame(
  state = c("PA", "GA", "NC", "MI", "WI", "NV", "AZ"),
  prob = c(0.5551, 0.6801, 0.1033, 0.5151, 0.3792, 0.7679, 0),
  votes = c(20, 16, 15, 16, 10, 6, 11)
)

# Function to calculate probability of a specific combination
calc_combo_prob <- function(combo, states) {
  prob <- 1
  for(i in 1:nrow(states)) {
    if(i %in% combo) {
      prob <- prob * states$prob[i]
    } else {
      prob <- prob * (1 - states$prob[i])
    }
  }
  return(prob)
}

# Generate all possible combinations
n_states <- nrow(states)
total_prob <- 0

# Check each possible combination
for(i in 1:(2^n_states - 1)) {
  # Convert number to binary to get combination
  combo <- which(intToBits(i)[1:n_states] == 1)

  # Calculate total electoral votes for this combination
  votes <- sum(states$votes[combo])

  # If this combination has enough votes, add its probability
  if(votes >= 44) {
    prob <- calc_combo_prob(combo, states)
    total_prob <- total_prob + prob
  }
}

# Trump
# Create data frame of states
states <- data.frame(
  state = c("PA", "GA", "NC", "MI", "WI", "NV","AZ"),
  prob = c(1-0.5551, 1-0.6801, 1-0.1033, 1-0.5151, 1-0.3792, 1-0.7679, 1-0), # Using 1-p for Trump's probabilities
  votes = c(20, 16, 15, 16, 10, 6, 11)
)

# Function to calculate probability of a specific combination
calc_combo_prob <- function(combo, states) {
  prob <- 1
  for(i in 1:nrow(states)) {
    if(i %in% combo) {
      prob <- prob * states$prob[i]
    } else {
      prob <- prob * (1 - states$prob[i])
    }
  }
  return(prob)
}

# Generate all possible combinations
n_states <- nrow(states)
total_prob <- 0

# Check each possible combination
for(i in 1:(2^n_states - 1)) {
  # Convert number to binary to get combination
  combo <- which(intToBits(i)[1:n_states] == 1)

  # Calculate total electoral votes for this combination
  votes <- sum(states$votes[combo])

  # If this combination has enough votes, add its probability
  if(votes >= 51) {
    prob <- calc_combo_prob(combo, states)
    total_prob <- total_prob + prob
  }
}

@@

```{r}
#| label: tbl-result
#| tbl-cap: Predicted Harris Win Percentage in Swing States
#| echo: false
#| message: false
#| warning: false

mean_probabilities %>%
  gt() %>%
  fmt_number(
    columns = vars(Harris_Win_Probability),
    decimals = 2
  ) %>%
  cols_label(
    State = "State",
    Harris_Win_Probability = "Harris Win Probability (%)",
    electoral_votes = "Electoral Votes"
  ) %>%
  tab_options(
    table.font.size = 12,
    heading.title.font.size = 16,
    heading.subtitle.font.size = 12,
    table.border.top.width = px(2),
    table.border.bottom.width = px(2)
  )

```
Combining with the fact that Harris requires 44 electoral votes to win and Trump requires 51, we have found using the probability function that Harris has a 38.3 chance to acquire more than 44 electoral votes, according to the probabilities and electoral votes in @tbl-result. On the other hand, Trump has a 61.7 probability to acquire more than 51 votes. Thus, our conclusion is that while both candidates possess a significant chance of winning, Trump has the greater probability of acquiring the necessary electoral votes from the seven swing states on election day.

## Popular vote

@fig-lradj presents the fitted values of the candidates' adjusted national support rates as well as the data points of polled support rates adjusted for historical pollster bias. It indicates that Trump is expected to have higher national support levels, but the two converges as election day approaches. This result is counter intuitive, as most national polls indicate that Harris will have the popular vote. However, the adjusted values for Trump's support level are significantly greater than the polled values because historically, the actual percentage of the popular vote for Trump in 2020 was significantly greater than his polled percentages of the popular vote. Thus, assuming that the 2024 national polls are equally biased against Trump compared to the 2020 polls, it is difficult to tell who will get the popular vote on election day. However, as previously stated, this assumption is likely false, as the candidates have changed.

```{r}
#| label: fig-lradj
#| fig-cap: Adjusted Popular Vote Prediction, 2024
#| echo: false
#| message: false
#| warning: false
# plot model predictions
# combine the datasets to create a single visual
combined_data_adj <- bind_rows(harris_2024_national_adj, trump_2024_national_adj)
combined_data <- bind_rows(harris_2024_national, trump_2024_national)

# create the plot that contains scattered points of adjusted percentages
# and lines for fitted values
ggplot(combined_data_adj, aes(x = end_date, y = pct_adj, color = candidate_name)) +
  # Scatter points for actual adjusted percentages
  geom_point(alpha = 0.6) +
  # Regression lines
  geom_line(aes(y = fitted_date), size = 0.5) +
  xlim(as.Date("2024-07-19"), as.Date("2024-10-25")) +
  # Assign specific colors
  scale_color_manual(values = 
                       c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  # Customize the plot appearance
  theme_classic() +
  labs(
    y = "Adjusted Percent Support",
    x = "End Date",
    color = "Candidate"
  ) +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

@fig-lr Presents the predicted unadjusted national support rates for Harris and Trump. The conclusions one can make from this plot is similar to those of most polls: Harris is expected to win the popular vote on any day, including the election day.

```{r}
#| label: fig-lr
#| fig-cap: Unadjusted Popular Vote Prediction, 2024
#| echo: false
#| message: false
#| warning: false
# Unadjusted support percentages
ggplot(combined_data, aes(x = end_date, y = pct, color = candidate_name)) +
  # Scatter points for actual adjusted percentages
  geom_point(alpha = 0.6) +
  # Regression lines
  geom_line(aes(y = fitted_date), size = 0.5) +
  xlim(as.Date("2024-07-19"), as.Date("2024-10-25")) +
  # Assign specific colors
  scale_color_manual(values = 
                       c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  # Customize the plot appearance
  theme_classic() +
  labs(
    y = "Adjusted Percent Support",
    x = "End Date",
    color = "Candidate"
  ) +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

```
The regression coefficients for the logistic and linear models are presented in the Appendix because they are not the outcome of interest. Please view @tbl-coe1 to @tbl-coe7.

@@

The dataset used for this analysis has a number of clear limitations. First of all, there is a lack of high-quality polls. Though FiveThirtyEight distributes each pollster with a rating out of 3.0, almost all pollsters under the rating of 2.7 produces outcomes that are unreliable and highly volatile. For this research, it was only possible for us to use pollsters with a 3.0 rating when analyzing national polls; we had to lower the standards to 2.5 for the analysis of swing states, as the number of high-quality state-specific polls are very limited. The inclusion of unreliable data reduces the predictability of our logistic models and thus our final outcome. The lack of predictability of our Nevada model is a perfect example of how low-quality data and the lack of data points impacts our ability to create effective models. However, we acknowledge that this is beyond the control of FiveThirtyEight, as they do not conduct the polls.

The second major challenge presented by the dataset is the lack of predictors. FiveThirtyEight utilizes predictive variables that are unavailable to us, such as weighted values that represent the extent to which a pollster is partisan [@5383]. The predictors used in this research, such as the transparency score of the pollster, are not related to their bias towards any candidate. This limits the predictive power of our models.

@@

## Diagnostics
```{r}
#| label: fig-pen
#| fig-cap: ROC Curve for Harris Win Prediction, Pennsylvania
#| echo: false
#| message: false
#| warning: false

@@

## Coefficients

```{r}
#| label: tbl-coe1
#| fig-cap: Logistic Regression Coefficients for Pennsylvania
#| echo: false
#| message: false
#| warning: false
# Tidy the logistic regression model
tidy_Pennsylvania <- tidy(logistic_Pennsylvania)

# Display the table using knitr::kable
kable(tidy_Pennsylvania, digits = 2, 
      col.names = c("Term", "Estimate", "Std. Error", "Statistic", "p-value"))

```
```{r}
#| label: tbl-coe2
#| fig-cap: Logistic Regression Coefficients for Georgia
#| echo: false
#| message: false
#| warning: false
# Tidy the logistic regression model
tidy_Georgia <- tidy(logistic_Georgia)

# Display the table using knitr::kable
kable(tidy_Georgia, digits = 2, 
      col.names = c("Term", "Estimate", "Std. Error", "Statistic", "p-value"))

```{r}
#| label: tbl-coe3
#| fig-cap: Logistic Regression Coefficients for Michigan
#| echo: false
#| message: false
#| warning: false
# Tidy the logistic regression model
tidy_Michigan <- tidy(logistic_Michigan)

# Display the table using knitr::kable
kable(tidy_Michigan, digits = 2, 
      col.names = c("Term", "Estimate", "Std. Error", "Statistic", "p-value"))

```{r}
#| label: tbl-coe4
#| fig-cap: Logistic Regression Coefficients for North_Carolina
#| echo: false
#| message: false
#| warning: false
# Tidy the logistic regression model
tidy_North_Carolina <- tidy(logistic_North_Carolina)

# Display the table using knitr::kable
kable(tidy_North_Carolina, digits = 2, 
      col.names = c("Term", "Estimate", "Std. Error", "Statistic", "p-value"))

```{r}
#| label: tbl-coe5
#| fig-cap: Logistic Regression Coefficients for Arizona
#| echo: false
#| message: false
#| warning: false
# Tidy the logistic regression model
tidy_Arizona <- tidy(logistic_Arizona)

# Display the table using knitr::kable
kable(tidy_Arizona, digits = 2, 
      col.names = c("Term", "Estimate", "Std. Error", "Statistic", "p-value"))

```{r}
#| label: tbl-coe6
#| fig-cap: Logistic Regression Coefficients for Wisconsin
#| echo: false
#| message: false
#| warning: false
# Tidy the logistic regression model
tidy_Wisconsin <- tidy(logistic_Wisconsin)

# Display the table using knitr::kable
kable(tidy_Wisconsin, digits = 2, 
      col.names = c("Term", "Estimate", "Std. Error", "Statistic", "p-value"))

```{r}
#| label: tbl-coe7
#| fig-cap: Logistic Regression Coefficients for Nevada
#| echo: false
#| message: false
#| warning: false
# Tidy the logistic regression model
tidy_Nevada <- tidy(logistic_Nevada)

# Display the table using knitr::kable
kable(tidy_Nevada, digits = 2, 
      col.names = c("Term", "Estimate", "Std. Error", "Statistic", "p-value"))
